# Introduction to R Objects  

At its core, R is an objected-oriented computational and programming environment. Everything in R is an object belonging to a certain *class*.  In this chapter we begin by discussing **atomic** objects, these are the core fundamental objects used in R.  Next we discuss vectors, lists, and matrices.  These objects are among the most common types that are used.  Lastly we discuss factors and data frames.  Factors and data frames are known as dynamic objects, and are more complex.  All of these objects have special properties and a multitude of features that we can use.  We discuss some of their key properties here, but will continue exploring and learning about their features, and introducing more object types throughout this document. 

## Atomic Objects 

R can represent different types of data. The types include `numeric`, `integer`, `complex`, `logical`, `character`, and `raw`. These are the basic fundamental objects we can use in R. For our class we will not need the `complex` type which stores complex numbers, and in practice `raw` is rarely used. We will concentrate on the remaining four types. Unlike other object-oriented languages we do not need to specify what type of object we are creating when we create it.  Instead, R guesses the type of object you are creating.  To check the object type we can use the  `class()` function.  function.  

### Numeric {-}

Numeric objects are perhaps the most common.  These are objects which contain a real number, that is, a number which can contain a decimal value.  These objects are comparable to `doubles` in `C`. 

```{r}
a = 17.45
a
class(a)
b = 5
b 
class(b)
```

Both the variables `a` and `b` are `numeric` objects. When you type a number R will default to treating it as a `numeric` object which allows decimals. 

### Integer {-}

We can also create numeric objects which are specifically made to store integer values.  We can do this using the `as.integer()` function.

```{r}
a = as.integer(a)
a
class(a)
b = as.integer(b)
b 
class(b)
```

### Logical {-}

Logical values are either `TRUE` or `FALSE` and are created by using logical and relational operators.  In other words, they are created by using statements that compare variables. There are several ways to do logical statements as we saw in Section \@ref(operators). 

```{r}
b = 5
n = (10<11)
n
class(n)
```

We can also assign a value as `TRUE` or `FALSE` manually by setting it equal to `TRUE` or `FALSE`, or by using `T` or `F`. 

```{r}
c = T
c
class(c)
```


### Character {-}

`Character` values are text. They are often used as data values and labels. 

```{r}
first = "George"
first 
class(first)
last = "Washington"
last
class(last)
```
There are several functions that can operate on character strings.

```{r}
full = paste(first, last)
full 
nchar(full)
tolower(full)
toupper(full)
```

The function `paste()` concatenates two or more character strings with a separator, which is a space by default. The function `nchar()` returns the number of characters in a string. The functions `tolower()` and `toupper()` changes any upper case characters to lower case and vice-versa.

## Vectors

All the objects we have created this far are single element *vectors*.  R is a vectorized language, meaning most of the procedures, functions, and operations have been optimized to work with vectors. It is typically advantageous to utilize this feature.  **A vector is a collection of values of the same data type**.  We can use the concatenate function, `c()`, to create vectors, and to make a vector larger. 

```{r}
v1 = c(19, 390.3, pi, -32.1)
v1
class(v1)

v2 = c(1.1, 6, -9.4, 32.1)
v2
class(v2)

```

If we try to create a vector with a mix of classes R will convert all the objects to be the same class.  In general, it is easiest to convert objects into a `character` but hard to convert `character` into something else.  Be cautious when mixing data types and vectors because you will not be notified if objects are converted, and they may not be converted to the class you intended. 

```{r}

v3 = c(v1, first)
class(v3)

v4 = c(first, last)
class(v4)
```


The `length()` function can be used to obtain the number of elements in a vector. 

```{r}
length(v1)
```

Vectors can be used in arithmetic computations. If the two vectors are of the same length, the computations are performed element-by-element.

```{r}
v1 + v2

v1 * v2
```

Single numbers (scalars) will operate on all the vector elements in an expression. 

```{r}
5*v1

v1/3
```

Individual elements of a vector can be obtained using an index in square brackets. A negative index removes that element from the vector. The `v2[-1]` is the vector `v2` with the first element removed. The concatenate function can be used to obtain two or more elements of a vector in any desired order. Here `v1[c(3,2)]` returns the third and second elements of the vector `v1`.

```{r}
v1[3]
v2[-1]
v3[c(3,2)]
```

## Lists 

Lists are thought of as a vector with a variety of classes.  A list is made up of elements, and each element can be of a different class. 

```{r}
lst = list(4, v4, v2)
lst
class(lst)
```

We can observe the class of each element in the list by using the `str()` function. 

```{r}
str(lst)
```

The above output tells us we have a list of three objects.  The first object is a numeric vector with one element, the second object is a character vector with two elements, and third object is a numeric vector with four elements. We can subset elements in a list using double brackets we `[[]]`.  Inside these square brakets we state the element we would like to obtain. 

```{r}
lst[[1]]
class(lst[[1]])
```

To determine how long our list is we can use the `length()` function.  

```{r}
length(lst)
```

 

## Matrices

A matrix is a two dimensional array of data of **the same type**. The matrix function, `matrix()`, can be used to create a new matrix.

```{r}
m = matrix(c(1, 9, 2, 0, 5, 7, 3, 8, 4), 
           nrow=3, ncol=3)
m
```

R labels the rows and columns for us in the output. The matrix is filled column-by-column using the elements of the vector created by the concatenate function. As with vectors, matrices can be used in arithmetic operations with scalars and other matrices of the same size.

```{r}
m2 = m/2
m2
m *m2
```

Indices can be used to obtain the elements of a matrix, but now we must consider both the row and column.

```{r}
m[2,2]
m[c(1,3), c(1,3)]
m[2,]
m[,3]
```

Some functions are particularly useful when using matrices.  For instance, `t()`, `dim()`, and `c()`. The transpose function, `t()`, switches the column and rows of a matrix. The dimension function, `dim()`, returns the dimensions (number of rows, columns) of a matrix. The concatenate function, `c()`, turns a matrix into a vector by concatenating the columns of the matrix.
```{r}

# Dimensions (row, column)
dim(m)

# Transpose
t(m)

# Convert to vector 
c(m)
```



## Factors

Factors are useful for categorical data. Factors differ from character objects in that a character object is a string of characters or symbols placed in a specific order.  For example, the object `first = "George"` is a character object with six elements.  In contrast, the collection of values "George" would instead represent a distinct value, or *level*, for a factor object.  We can create a factor object using the `factor()` function. 

```{r}
colors = c("red", "blue", "red", "red", "blue")
colors = factor(colors)
colors
class(colors)
```

Here the unique elements in the factor are called "levels".  There are only two levels `red` and `blue`, and there are five elements in our factor object. We can index factors in the same way we index a list.  

```{r}
# Return first element in the factor object 
colors[1]
```


## Data Frames

Like a matrix, a data frame is a rectangular array of values where each column is a vector, However, unlike a matrix, the columns can be different data types. 

We can create a set of vectors of the same length and use the `data.frame()` function to make a data frame object.

```{r}
age = c(1, 8, 10, 30, 31)
gender = c("Female", "Female", "Male","Female","Male")
married = c(FALSE, FALSE, FALSE, TRUE, TRUE)
simpsons = data.frame(age, gender, married)
simpsons
class(simpsons)
```


Data frames can be indexed using the same techniques as was done with matrices. 

## Other Object Types and the Global Environment

There are more objects then what we have discussed above. For example, many of the advanced functions create specific objects generated by that specific function.  There are hundreds, and possibly thousands, of such objects.  These objects generally are special cases of lists, factors, and other various types of objects that we have defined in this section.  The objects we have described here are the building blocks of most values we will be working with. Functions like `class()` and `length()` are also considered as objects, but are of a different type.  We discuss functions in more detail in section \@ref(functions).  

There are also built-in, or special objects in R.  For example, the object `pi` is an object already defined. These built-in values and functions can be written over, but that is not advised.

```{r}
pi
```

Every time we create an object we see that the Global Environment tab in the top right pane updates.  The object we have created is now listed in the Global Environment. This is a collection of all *user created* objects in R, that R knows about, and that R can easily call.  Built-in objects, such as `pi`, will not be listed here. 


## Additional Resources 

- Chapters 2, 3, 4.1, 4.3, 5.1-5.3, 6 of [CRAN Intro-to-R Manual](https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf)
- Videos: 
    + [Variables 1 | Types and Assignments](https://ucr.yuja.com/V/Video?v=2368642&node=8487537&a=1529691043&autoplay=1)
    + [Variables 2 | Nameing Conventions and Best Practices](https://ucr.yuja.com/V/Video?v=2368641&node=8487536&a=957339369&autoplay=1)
    + [Vectors 1 | Introduction](https://ucr.yuja.com/V/Video?v=2368859&node=8488053&a=283774152&autoplay=1)
    + [Vectors 2 | Subsetting and Modifying](https://ucr.yuja.com/V/Video?v=2368857&node=8488051&a=1465899289&autoplay=1)
    + [Vectors 3 | Vectorized Functions - Logical Comparisons](https://ucr.yuja.com/V/Video?v=2368856&node=8488050&a=612212822&autoplay=1)
    + [Matrices 1 | Introduction](https://ucr.yuja.com/V/Video?v=2368855&node=8488049&a=447596964&autoplay=1)
    + [Matrices 2 | Accessing Rows and Columns](https://ucr.yuja.com/V/Video?v=2368854&node=8488047&a=1529352108&autoplay=1)


# More on R Objects 

Factors and lists have especially unique properties that are often utilized.  In this chapter we take special care to discuss these properties. 

## Factors 

In real-world problems, you often encounter data that can be classified in categories. For example, suppose a survey was conducted of a group of seven individuals, who were asked to identify their hair color and if they own a pet.

```{r}
name = c("Amy", "Bob", "Eve", "Kim", "Max", "Ray", "Sam")
hair = c("Blonde", "Black", "Black", "Red", "Blonde", "Brown", "Black")
own_pets = c(TRUE, FALSE, TRUE, TRUE, FALSE, FALSE, FALSE)

catagorical = data.frame(name, hair, own_pets)
colnames(catagorical) = c("Name", "Hair Color", "Own Pets")

catagorical

```


Here, the hair color and pet ownership are examples of categorical data. For the hair color variable we will typically want to store it as a factor, as opposed to a character vector. The different values that the factor can take are called levels. In R, you can create a factor with the `factor()` function. 

```{r}
f = factor(hair)
f
```

A factor looks like a vector, but it has special properties. Levels are one of them. Notice that when you print the factor, R displays the distinct levels below the factor. R keeps track of all the possible values in a vector, and each value is called a level of the associated factor.The `levels()` function shows all the levels from a factor.

```{r}
levels(f)
```


If your vector contains only a subset of all the possible levels, then R will have an incomplete picture of the possible levels. Consider the following example of a vector consisting of directions:

```{r}
directions = c("North", "West", "North", "East", "North", "West", "East")
f = factor(directions)
f
```

Notice that the levels of your new factor do not contain the value “South”. So, R thinks that North, West, and East are the only possible levels. However, in practice, it makes sense to have all the possible directions as levels of your factor. To add all the possible levels explicitly, you specify the `levels` argument of the function `factor()`.

```{r}
directions = c("North", "West", "North", "East", "North", "West", "East")
f = factor(directions,
            levels = c("North", "East", "South", "West"))
f
```

R lets you assign abbreviated names for the levels. You can do this by specifying the `labels` argument of `factor()`.

```{r}
directions = c("North", "West", "South", "East", "West", "North", "South")
f = factor(directions,
            levels = c("North", "East", "South", "West"),
            labels = c("N", "E", "S", "W"))
f
```

Sometimes data has some kind of natural order between elements. For example, sports analysts use a three-point scale to determine how well a sports team is competing: 

**loss < tie < win**.

In market research, it’s very common to use a five point scale to measure perceptions: 

**strongly disagree < disagree < neutral < agree < strongly agree**.

Such kind of data that is possible to place in order or scale is known as **ordinal data**. We can store ordinal data as an ordered factor. To create an ordered factor, use the `factor()` function with the argument `ordered=TRUE`.  

```{r}
record = c("win", "tie", "loss", "tie", "loss", "win", "win")
f = factor(record, 
            ordered = TRUE)
f
```


You can also reverse the order of levels using the `rev()` function. 

```{r}
record = c("win", "tie", "loss", "tie", "loss", "win", "win")
f = factor(record, 
            ordered = TRUE, 
            levels = rev(levels(f)))
f

```


If you have no observations in one of the levels, you can drop it using the `droplevels()` function.

```{r}
record = c("win", "loss", "loss", "win", "loss", "win")
f = factor(record,
            levels = c("loss", "tie", "win"))

f

droplevels(f)

```

The `summary()` function will give you a quick overview of the contents of a factor.


```{r}
f = factor(hair)
f
summary(f)
```

The function `table()` tabulates observations.


```{r}
table(f)
```



## Lists 

A *list* is an array of objects. Unlike vectors and matrices, the elements in a list can belong to different classes. Lists are useful for packaging together a set of related objects. We can create a list of objects in our environment by using the `list()` function.

```{r}
lst = list(1, 2, 3)

# A list of characters
lst = list("red", "green", "blue")

# A list of mixed datatypes
lst = list(1, "abc", 1.23, TRUE)
```

The best way to understand the contents of a list is to use the structure function `str()`. It provides a compact display of the internal structure of a list.

```{r}
lst = list(1, "abc", 1.23, TRUE)
str(lst)
```


A list can contain sublists, which in turn can contain sublists themselves, and so on. This is known as *nested list* or *recursive vectors*.

```{r}
lst = list(1, 3, "abc", list("a","b","c"), TRUE)
str(lst)
```

There are two ways to extract elements from a list:

- Using `[[]]` gives you the element itself.
- Using `[]` gives you a list with the selected elements

You can use `[]` to extract either a single element or multiple elements from a list. However, the result will always be a list.

```{r}
# extract 2nd element
lst[2]

# extract 5th element
lst[5]

# select 1st, 3rd and 5th element
lst[c(1,3,5)]

# exclude 1st, 3rd and 5th element
lst[c(-1,-3,-5)]

```


You can use `[[]]` to extract only a single element from a list. Unlike `[]`, `[[]]` gives you the element itself.

```{r}
# extract 2nd element
lst[[2]]

# extract 5th element
lst[[5]]

```

You can’t use logical vectors or negative numbers as indices when using `[[]]`.  The difference between `[]` and `[[]]` is really important for lists, because `[[]]` returns the element itself while `[]` returns a list with the selected elements. The difference becomes clear when we inspect the structure of the output – one is a character and the other one is a list.

```{r}
lst = list("a","b","c","d","e","f")

class(lst[[1]])

class(lst[1])
```

Each list element can have a name. You can access individual element by specifying its name in double square brackets `[[]]` or use `$` operator.

```{r}
months = list(JAN=1, FEB=2, MAR=3, APR=4)

# extract element by its name
months[["MAR"]]


# same as above but using the $ operator
months$MAR


# extract multiple elements
months[c("JAN","APR")]


```

You can access individual items in a nested list by using the combination of `[[]]` or `$` operator and the `[]` operator.

```{r}
lst = list(item1 = 3.14,
            item2 = list(item2a = 5:10,
                         item2b = c("a","b","c")))

# preserve the output as a list
lst[[2]][1]


# same as above but simplify the output
lst[[2]][[1]]


# same as above with names
lst[["item2"]][["item2a"]]


# same as above with $ operator
lst$item2$item2a


# extract individual element
lst[[2]][[2]][3]


```

Modifying a list element is pretty straightforward. You use either the `[[]]` or the `$` to access that element, and simply assign a new value.

```{r}
# Modify 3rd list element
lst = list("a","b","c","d","e","f")
lst[[3]] = 1
str(lst)

```

You can modify components using `[]` as well, but you have to assign a list of components.

```{r}
# Modify 3rd list element using []
lst = list("a","b","c","d","e","f")
lst[3] = list(1)
str(lst)

```


Using `[]` allows you to modify more than one component at once.

```{r}
# Modify first three list elements
lst = list("a","b","c","d","e","f")
lst[1:3] = list(1,2,3)
str(lst)

```

You can use same method for modifying elements and adding new one. If the element is already present in the list, it is updated else, a new element is added to the list.

```{r}
# Add elements to a list
lst = list(1, 2, 3)
lst[[4]] = 4
str(lst)


```

By using `append()` method you can append one or more elements to the list.

```{r}
# Add more than one element to a list
lst = list(1, 2, 3)
lst = append(lst,c("a","b","c"))
str(lst)


```


By using a logical vector, you can remove list elements based on the condition.

```{r}
# Remove all negative list elements
lst = list(-4,-3,-2,-1,0,1,2,3,4)
lst[lst <= 0] = NULL
str(lst)
```


The `c()` does a lot more than just creating vectors. It can be used to combine lists into a new list as well.

```{r}
lst1 = list("a","b","c")
lst2 = list(1,2,3)
lst = c(lst1, lst2)
str(lst)
```

Basic statistical functions work on vectors but not on lists. For example, you cannot directly compute the mean of list of numbers. In that case, you have to flatten the list into a vector using `unlist()` first and then compute the mean of the result.

```{r}
lst = list(5, 10, 15, 20, 25)
mean(unlist(lst))
```

To find the length of a list, use `length()` function.


```{r}
length(lst)
```

