# Working with Data Sets

In this section we discuss different methods for loading data sets into our R session.  There are many different files we can create and import.  We will focus our attention on loading csv files because they tend to be easier to import, and they are one of the more typical file types that are used. In the second half of the this chapter we introduce basic data manipulation strategies and helpful functions when working with data sets. 


## Getting Data Sets in Our Working Environment

### Built-In Data {-}

As discussed last week, there are built in objects which are not loaded into the global environment, but can be called upon at any time.  For example `pi` returns the value `r pi`.  Similarly, there are built in data sets that are ready to be used and loaded at a moments notice.

1) To see a list of built in data sets type in the console: 

```{r}
data()
```

2) These data sets can be used even if they are not listed in the global environment. For example, if you would like to load the data set \texttt{cars} in the global environment, run the following command: 

```{r, eval = F}
data("cars")
```


### Importing From Your Computer{-}

Although built-in data sets are convient, most the time we need to load our own datasets. We load our own data sets by using a function specifically designed for the file type of interest.  This function usually uses the file path location as an argument. This can be done in many different ways; however, we will only go over two. 

**Option 1**

1) Download the file `InsectData.csv` from **ELearn**.  Save this file in a spot in your computer you will remember. 

2) In the **Environment** window (upper left window), click on the **Import Dataset** button.  A drop down menu will appear.  Select the **From Text (base)...** option.  Find the file `InsectData.csv` and select it. 

3) A pop up menu will appear giving you options for loading in the file, and showing a preview of what the file will look like once loaded. Select the appropriate options and click **Import**.

4) A new line of code has generated in the console which will read the data into your current environment.  Copy and paste this into your R script document if you would like to save this line of code for later.  You will have to reload this file into your environment each time you start a new R session and would like to use this file. 

**Option 2**

1) Download the file `InsectData.csv` from **ELearn**.  Save this file in a spot in your computer you will remember. 

2) In the lower right hand window select the **File** tab.  Now search for the file which you have saved `InsectData.csv`.  

3) Click on the file `InsectData.csv` in order to see a dropdown menu.  Select **Import Dataset...**

4) A window will appear which will give you options and a preview of your file. Select appropriate options if needed then click **Import**. 

5) A new line of code has generated in the console which will read the data into your current environment.  Copy and paste this into your R script document if you would like to save this line of code for later.  You will have to reload this file into your environment each time you start a new R session and would like to use this file. 


### Import From Online {-}

We can also download data sets from online in a variety of different ways. Below is one option.  With this method we are using the same `InsectData.csv` file, but it has been posted online.  We feed the url of where the data set has been posted into the `read.csv()` function in order to open the file. 

```{r}

the_url = "https://raw.githubusercontent.com/rpkgarcia/LearnRBook/main/data_sets/InsectData.csv"
the_data = read.csv(the_url)
```


## Basic Data Manipulation

Lets recall a few useful things about data frames. As we learned already, data sets are contained in an object called a data frame. One can view this as a specialized table or matrix of rows and columns, where each column is a data variable, such as height or age, and each row is a single observation. All of the values within a column must be the same data type (numeric,factor, logical, etc.). Data frames can be created or called within R, imported from text or spreadsheet files, or imported from the web. 

```{r}
group = c("G1", "G2", "G1", "G1", "G2")
age = c(35, 30, 31, 28, 40)
height = c(65, 70, 60, 72, 68)
pets = c(TRUE, TRUE, FALSE, FALSE, TRUE)

mydata = data.frame(group, age, height, pets)
mydata
```


We have various R functions that help provide information about the data frame.

```{r}
# How many rows
nrow(mydata)

# How many columns
ncol(mydata)

# Rows, Columns
dim(mydata)

# Colnames
colnames(mydata)
```


The `summary()` function is a powerful command that gives you some summary statistics about the variales in the data frame. 

```{r}
summary(mydata)
```

The summary statistics are listed below the names of the variables. Since pets is a logical variable, R gives you the frequencies of each unique value. In this example there are three values of `TRUE` and two values of  `FALSE`. Since age and weight are numeric, R computes and returns the minimum, 1st quartile (25th percentile), median, mean, 3rd quartile (75th percentile), and maximum values. If you have many data values, this is a quick way to get a feel for how the data are distributed.


The `table()` function can also be used to cross-tabulate categorical data.  Let’s create a frequency table for the different groups.

```{r}
table(mydata$group)
```

 We can also create a frequency table of pet status for both groups. 

```{r}
table(mydata$group, mydata$pets)
```

### Indexing {-}

As with matrices, square brackets followed by the row, column can be used to return a specific data value. For example, to get the datum in the 3rd row, 2nd column.

```{r}
mydata[3,2]
```

If you leave out the number before or after the comma, the entire column or row is returned. 

```{r}
mydata[3,]

mydata[,2]

```

For data frame you can also access variables by name instead of column number. Use a dollar sign and the name of the variable (no spaces). This usually makes the R code more readable. 

```{r}
mydata$height[4]

mydata$height
```

The colon operator can be used to generate a sequence of indices. Typing a:b will produce a range of integers $a$, $a+1$,..., $b$.

```{r}
mydata[2:4,]

mydata$age[3:5]
```

You can use logical conditions to obtain subsets of the data. A logical value of `TRUE` will return a data row and a logical of `FALSE` omits that row.

```{r}
index_keep = (mydata$group =="G1")
group1 = mydata[index_keep, ]
group1
```

This operation can also be accomplished by using the `subset()` function. 

```{r}
group1 = subset(mydata, mydata$group == "G1")
group1
```


To subset by all values which are NOT equal to a condition we can use the logical operator `!=`.
```{r}
group2 = subset(mydata, mydata$group != "G1")
group2
```

### Rearranging Rows {-}

Data can be sorted using the `order()` function. This function returns the ranks of the variable being sorted. Including more than one variable allows a "nested sort," where the second variable, third variable, etc., is used when there are ties in the sorting based on the previous variables.

```{r}
order(mydata$height)

mydata[order(mydata$height),]
```

The ranks are used as row indices for the data frame. Notice that the observations are sorted by the last column, `weight`. We can sort by more than one variable. Let’s first sort by `group` alone and then by `group` followed by `weight` and see what we get.


```{r}
mydata[order(mydata$group), ]

mydata[order(mydata$group, mydata$height), ]

```

To reorder a vector from smallest to largest we can also condiser the sort() function. 

```{r}
sort(mydata$age)
```

### Adding Columns {-}

One can add a new variable (column) to a data frame by defining a new variable and assigning values to it. Below we add a `weight` variable to the data frame.

```{r}
wghts = c(169, 161, 149, 165, 155)
wghts

mydata$weight = wghts
mydata
```

We can also add a new column using the `cbind()` function. 

```{r}
mydata = cbind(mydata, wghts)
mydata
```

### NA values {-}

In addition, if we have a missing value, or a blank value, we can use the object `NA` to indicate the lack of a value. 

```{r}
fav_color = c("Red", NA, "Purple", NA, "Red")

mydata = cbind(mydata, fav_color)
mydata
```

We can drop check for `NA` values using the `is.na()` function. 

```{r}
is.na(mydata)
```

We can remove all rows with `NA` values using `na.omit()`. 
```{r}
na.omit(mydata)
```


### NULL {-}

One can drop a variable (column) by setting it equal to the `R` value `NULL`.

```{r}
mydata$wghts = NULL
mydata
```

Be careful using these methods. Once a variable or row is dropped, it’s gone. 

### Adding Rows {-}

Rows can be added to a data frame using the `rbind()` (row bind) function. Because our columns have different data types, we will create a list object and then add it as a new row. 


```{r}
newobs = list("G1", 23, 62, FALSE, 160, "Blue")
newdata = rbind(mydata, newobs)
newdata
```

We can also use `rbind()` to append one data frame to another. We can do this with the variables `group1` and `group2` created above still exist in your R environment. 

```{r}
group1
group2
rbind(group1, group2)
```

We can also remove rows from a data frame. Using negative index values has the same effect as including all the indices except the negative ones.

```{r}
mydata[-c(2,4), ]
```



