# If Statements


## The `if else` chain

In R, you can tell your code to run ONLY if a certain statement is true by making a `if else` chain.  For example;
```{r}
Do = "Add"
X = 2
Y = 4
if(Do == "Add"){
  Z = X+Y
}
Z
```
Only if the variable `Do` is equal to the character string `"Add"` will the variable `Z` be created.  Otherwise, whatever is in the body of the `if` statement is not run.
 
Here is an example of a silly function with an `if else` chain:
```{r}
Toyfun = function(X,Y,Do){
  if(Do == "Add"){
    Z = X+Y

  }else if(Do =="Subtract"){
    Z = X-Y

  }else if(Do =="Multiply"){
    Z = X*Y

    
  }else if(Do =="Penguin"){
    Z = c("<('' )")
    
  } else{
    Z = c(X,Y)
  }
  
  return(Z)
}
Toyfun(2,4,"Add")
Toyfun(2,4,"Subtract")
Toyfun(2,4,"Penguin")
Toyfun(2,4, "Typo")
```

A few notes:

- `if` will be written initially in the if statement code.

- `else` if will be executed if code if is not true.

- `else` will be executed if none of them are true.


## The `ifelse` function
There is a simple function that can check a logical argument or vector, and give back two answers; one if the statement is
true, and another if the statement is false. This function is `ifelse()`. For a simple example:
```{r}
Pet = "Dog"
ifelse(Pet == "Dog", "Woof","???")
Pet = "Cat"
ifelse(Pet == "Dog", "Woof","???")
```

We can also pass `ifelse()` a vector, and it will check the logical condition for each element of the vector.  For example, in the `USArrests` data, if I want to find the proportion of states who had a murder rate of over 10 per 100,000 and an urban population of 
over 50%:

```{r}
dangerous = ifelse(USArrests$Murder > 10 & USArrests$UrbanPop > 50, TRUE, FALSE)
sum(dangerous)/length(dangerous)
```
