# Packages 

`R` packages are a collection of `R` functions, complied code and sample data. They are stored under a directory called "library" in the `R` environment. By default, `R` installs a set of packages during installation. More packages are added later, when they are needed for some specific purpose. When we start the `R` console, only the default packages are available by default. Other packages which are already installed have to be loaded explicitly to be used by the `R` program that is going to use them.

All the packages available in `R` language are listed at [R Packages](https://cran.r-project.org/web/packages/available_packages_by_name.html).

To see a list of all packages installed on your device. 

```{r}
library()
```


To see a list of all packages that are currently loaded. 

```{r}
search()
```

When adding a new package to our library we only have to install it once. We can do so with the following command.

```{r, eval = F}
install.packages("library name")
```

Alternatively, we can also go to the lower left hand window and select the *Packages* tab.  Then hit the button **Install**.  A dropdown menu will appear where we can search for the package name. 

Before a package can be used in the code, it must be loaded to the current R environment. You also need to load a package that is already installed previously but not available in the current environment.

```{r, eval = F}
library("library name")
```



For example, suppose we wanted to install and load the package "ggplot2", (a very popular package for making plots).  We would type the following commands. 


```{r, eval = F}
# Install package (only need to do this once)
install.packages("ggplot2")

# Load into working environment (need to do this for each new R session)
library("ggplot2")
```



```{r, eval = F,echo=F}

### Resources


# I did use these resources. 
- https://www.dummies.com/programming/r/how-to-use-the-r-help-files/

```

